---
format:
  gfm:
    default-image-extension: ""
---

<!-- README.md is generated from README.qmd. Please edit that file -->

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# grumpy

<!-- badges: start -->
[![R-CMD-check](https://github.com/Bisaloo/grumpy/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Bisaloo/grumpy/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/Bisaloo/grumpy/graph/badge.svg)](https://app.codecov.io/gh/Bisaloo/grumpy)
<!-- badges: end -->

The `{grumpy}` R package provides a way to read NumPy's `.npy` files into R. It supports a wide range of data types and array shapes.

As a file format generated by a Python package, `.npy` files are prime candidates for using the `{reticulate}` package to read them into R.
However, this comes with downsides in terms of performance, flexibility, and robustness of the R package infrastructure. 
`{grumpy}`, on the other hand, is a pure R package with a single dependency (`{jsonlite}`), and is performant, flexible. Overall, it is designed to be used deep in the dependency graph of other packages.

For more details on the motivation and design principles underpinning `{grumpy}`, see the dedicated vignette: `vignette("design", package = "grumpy")`.

## Installation

You can install the released version of grumpy with:

``` r
install.packages("grumpy")
```

and the development version like so:

``` r
# install.packages("pak")
pak::pak("Bisaloo/grumpy")
```

## Example

```{r}
library(grumpy)
read_npy(system.file("extdata", "test_2d.npy", package = "grumpy"))
```
