---
title: "Beyond grumpy and `.npy` files"
vignette: >
  %\VignetteIndexEntry{Beyond grumpy and `.npy` files}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

In the other vignettes, we mentioned that whenever possible, we would recommend users to use more advanced and performant formats such as [Zarr](https://zarr-specs.readthedocs.io/en/latest/v3/core/). Zarr is particularly well-suited for `array`s or `matrix`es in R.

In this short vignette, we show how one would write and read Zarr files in R, using the `{Rarr}` package.

We also quickly and informally benchmark the performance of these formats against `.npy` files, using the `{grumpy}` package.

```{r setup}
library(grumpy)
library(Rarr)
```

## Introduction to Zarr

If you now understand how the `.npy` file format works, the Zarr format is a natural extension of it.
Zarr is a chunked, compressed, N-dimensional array storage format. Each chunk of data corresponds to a small sub-array of the original array, and is stored in a separate file, **in a format nearly identical to `.npy`**. The chunk is then usually compressed with high-performance compression algorithms such as `zlib`, `blosc`, or `zstd`.

One benefit of this chunked approach is that it allows for efficient reading and writing of large arrays, as only the necessary chunks need to be read or written, rather than the entire array.
This is particularly useful for large datasets that do not fit into memory, as it allows for out-of-core processing of the data.
It also provides strong benefits when some chunks are "empty" (e.g. filled with zeros, such as the white/black background of an image), as these chunks can be skipped during reading and writing, and thus reduce the overall storage requirements of the dataset.

```{r}
x <- matrix(0L, nrow = 1000L, ncol = 1000L)
x[300L:400L, 700L:800L] <- seq_len(101L * 101L)

f_zarr <- withr::local_tempfile(fileext = ".zarr")

Rarr::write_zarr_array(x, f_zarr, chunk_dim = c(100L, 100L), compressor = NULL)
```

## Storage size comparison

An equivalent `.npy` is 800 MB (`10 000 * 10 000 * 8` bytes). Let's look at the size of the Zarr file on disk:

```{r}
size <- list.files(f_zarr, full.names = TRUE, recursive = TRUE) |>
  file.info() |>
  subset(select = "size") |>
  sum()
size
```

**Without compression**, the equivalent Zarr data is thus 320 kB on disk, so `r round(8e8 / size)` times smaller than the `.npy` file. We could also use compression to further reduce the size of the Zarr file on disk, but this is out of scope for this vignette.

## Decoding speed comparison

We can also compare the speed of reading the `.npy` file with `{grumpy}` and the Zarr file with `{Rarr}`. To do so, we need to actually generate the data. We do it via `{reticulate}` because writing `.npy` files is out of scope for `{grumpy}`.

```{r}
library(reticulate)
np <- import("numpy")

f_npy <- withr::local_tempfile(fileext = ".npy")
np$save(f_npy, x)
```

```{r}
bm <- bench::mark(
  grumpy = read_npy(f_npy),
  zarr = read_zarr_array(f_zarr),
  iterations = 50L
)
bm
summary(bm, relative = TRUE)
```

There is a small time penalty for reading the Zarr file, since the various chunks need to be read and concatenated together, but the memory footprint is much smaller, and the Zarr file is much smaller on disk. This is particularly important for large datasets that do not fit into memory, as it allows for out-of-core processing of the data.
